The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-28 VegOzone.tex
Suggested changes by Frode.

{\bf TODO: Frode/Terje, you may help elaborate on the following paragraph on weather systems. I'm not an expert on synoptic scale meteorology and basically translated from Norwegian/Swedish into English.}

Following a report by the Norwegian Meteorological institute ({\bf TODO: How to cite report of Met NOR?}), the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark) and northwest of Norway, it was amongst the warmest and driest on record - favorable conditions for boundary layer ozone formation.

An unusually weak and northward shifted jet stream allowed for a persistent high pressure system above Fennoscandia which blocked the east-bound low pressure systems for several consecutive months. While in May and July whole Fennoscandia saw exceptionally warm and dry weather, the high pressure system's influence subsided in northern Fennoscandia in June allowing for lower temperatures than in the south.

The such induced heat and drought stress on vegetation was exceptional not only in large parts of Fennoscandia but also in the rest of Europe, where the influence of the high pressure system extended even over a period of five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors (\chem{CO}, \chem{VOCs}).

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 ({\bf TODO: Cite https://www.dsb.no/reportasjearkiv/brannstatistikk-2018/}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history ({\bf TODO: How to cite governmental report of Sweden? https://www.regeringen.se/rattsliga-dokument/statens-offentliga-utredningar/2019/02/sou-20197/ and report therein}).\\

>>>>

Following a report by the Norwegian Meteorological institute ({\bf TODO: Citation https://www.met.no/publikasjoner/met-info/met-info-2019}), the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record - favorable conditions for boundary layer ozone formation.

An unusually weak and northward shifted jet stream allowed for a persistent high pressure system above northern Europe, including Fennoscandia which blocked the pressure systems for several consecutive months. In the periode May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. Whgile southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whle had close to normalprecipitation, but with local variations.

Heat and drought stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high pressure system extended even over a period of five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors (\chem{CO}, \chem{VOCs}).

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 ({\bf TODO: Cite https://www.dsb.no/reportasjearkiv/brannstatistikk-2018/}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history ({\bf TODO: How to cite governmental report of Sweden? https://www.regeringen.se/rattsliga-dokument/statens-offentliga-utredningar/2019/02/sou-20197/ and report therein}).

Elevated ozone in Europe was liekly due to several other factors as well, like in the 2003 drought period, such as biogenic VOC emissions, reduced dry deposition, reduced clouds and increased solar radiation, photochemical reactions, and stagnant weather conditions \citep{JGR:Solberg2018}\\
